# Geometry-MMALS G1 v1.0.8
## Revised Direct Context Geometry and Global Fiber Alignment

**Build revision:** `reviewer-loss-design-r2`  
**Scientific status:** C0 protocol implementation only.

This revision incorporates the reviewer discussion on direct context geometry. The functional context is L2-normalized and the same normalized representation is used by the context-bottleneck router, geometry losses, metrics, and causal interventions. The raw context is retained only for audit diagnostics.

## Core mathematical design

For raw context $c^{raw}_{s,a}$:

$$
c_{s,a}=\frac{c^{raw}_{s,a}}{\max(\|c^{raw}_{s,a}\|_2,\varepsilon)}.
$$

The normalized half-chord distance is

$$
d_C(c_{s,a},c_{s,b})=\frac{1}{2}\|c_{s,a}-c_{s,b}\|_2.
$$

For the factor span $U_{max}=120^\circ$:

$$
\tau(u_a,u_b)=\sin\left(\frac{\pi}{2}\min\left(\frac{|u_a-u_b|}{U_{max}},1\right)\right).
$$

The protocol separates:

- local context matching;
- far-pair separation;
- per-source path-spread anti-collapse;
- interval-wise cross-source fiber alignment;
- factor-centroid grounding;
- route geometry;
- retention and host-function terms.

The full objective is

$$
\mathcal L=\mathcal L_{CE}+\lambda_{anchor}\mathcal L_{anchor}+\lambda_r\mathcal L_{route}+\lambda_c\mathcal L_{context}+\lambda_f\mathcal L_{far}+\lambda_v\mathcal L_{spread}+\lambda_a\mathcal L_{fiber}+\lambda_g\mathcal L_{centroid}+\lambda_h\mathcal L_{host-div}.
$$

## Primary ablations

1. no geometry;
2. route geometry only;
3. context geometry plus spread;
4. route plus context geometry;
5. full local and global alignment;
6. dimension-4 full alignment as a capacity probe.

Dimension 2 is the primary geometric test. Dimension 4 is not treated as a parameter-matched causal contrast.

## Evidence and falsification

The notebook exports source-block correlation and stress, centroid order, fiber resultant length, held-out factor decoding, path variance, effective rank, raw norm, near/far distance separation, accuracy, NLL, forgetting, and source-disjoint causal evidence.

A local context-geometry candidate requires a positive paired-bootstrap delta in source-level context rho and a negative stress delta. A global alignment candidate additionally requires improved centroid order, fiber resultant length, held-out decoding, and non-collapse diagnostics.

No single seed promotes a qualification claim.
